---
title: "3k running assignment"
author: "Ciaran Daly, id=21381943"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
 word_document: default
 # html_document: default
 # if you don't have Microsoft Word, you can output an html document and save it as a pdf for submission.
 # just uncomment (remove the #) from the "# html_document: default" line and add a # in front of the "word_document: default" line
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Inference and linear regression for 3 km Running Times based on laboratory testing

### Study Description:

Sixteen well-trained male middle and long distance runners performed independently a 3 km time trial and a number of running tests in the laboratory.  

One of the tests was running velocity at a blood lactate concentration of 4 mmol/L (`v4mM`), measured in km/hr.  For background knowledge to understand something about the study, it requires considerable physical effort to achieve a blood lactate concentration of 4 mmol/L in well-trained athletes (if you want to know more about blood lactate and athletic performance here is a reference https://www.trainingpeaks.com/blog/what-is-lactate-and-lactate-threshold/).  

Other tests were running velocity at Lactate Threshold (`vTlac`), and at VO2 max (`V02Max`) both measured in km/hr.  In addition, a measure of running economy, oxygen uptake (measured in ml/kg/min) at running speeds of 14.5km/hr and 16.1 km/hr were measured and the variable names are `Rel.14.5` and `Rel.16.1`.  

All the laboratory testing took place on a motorised treadmill, and distance running performance was determined by 3 km time trials on an indoor 200m track, 3k time in minutes was recorded as `Running.Time`.

Based on data on other populations, the sports scientist doing the study wants to know what evidence the sample provides that the mean 3k running time in this population is different to 10 minutes 12 seconds (10.2 minutes), and also wants to estimate a plausible interval for the population mean 3k running time.  The sports scientist also wants to estimate the 3k time below which the fastest 10% of the population times lie.  Finally, the sports scientist wants to investigate if it is possible to predict 3k running time from running velocity at a blood lactate concentration of 4 mmol/L.

### Aims: 

To investigate the following:

(1) Is the population mean 3k running time different to 10 minutes 12 seconds (ie as decimal 10.2 minutes)?  What is a plausible range in which the population mean 3k time is likely to lie?  And what time are we confident that the top 10% of the population 3k times will be faster than?

(2) Can we use linear regression to predict 3 km running time in minutes (`Running.Time`) from running speed at lactate 4 mmol/L in km per hour (`v4mM`)?

It is decided in advance that the acceptable significance level for hypothesis testing is $\alpha$ = 0.05.

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# load required libraries
library(tidyverse)
library(infer)
library(tolerance)
```

### Read the data and see a few rows

```{r}
running = read.csv("3krunning.csv", header = TRUE)
head(running)
```

### Summary Statistics

Shown below are summary statistics for some of the columns in the dataset. 

```{r}
running %>% summarise(
  count = n(),
  meantime = mean(Running.Time),
  sdtime = sd(Running.Time),
  meanv4mM = mean(v4mM),
  sdv4mM = sd(v4mM),
  meanRel14.5 = mean(Rel.14.5),
  sdRel14.5 = sd(Rel.14.5),
  
       )

```


## Hypothesis test and confidence interval for mean 3k time

#### Task 1: Write the null and alternative hypotheses and significance level of the test (text, 1 sentence)
i.e. the hypothesis test for Aim (1), first sentence.

Null Hypothesis H0: The population mean 3k running time = 10.2 minutes

Alternative Hypothesis H1: The population mean 3k running time ≠ 10.2 minutes, at a 0.05 significance level
 
**5 marks**

#### Task 2): Appropriate statistical test and assumptions

(a) What statistical test is appropriate to test the Task 1 hypotheses?  State the assumptions needed for the test to be valid, and how they could be justified in this study. (1-2 sentences).  

To test the Task 1 hypotheses, a t-test would be appropriate, as we can compare a sample mean to the true mean (10.2 minutes). We can assume that the data is normally distributed and that the sample population is randomly selected.


(b) Comment on the output of the R code below in relation to an assumption for the test (1 sentence)

Based on the output below, we can assume the data is normally distributed, and since the the mean and median are equal in data that is normally distributed, it can be assumed that the mean may not be 10.2 minutes and closer to 9.4 minutes, which indicates we may be rejecting the null hypothesis.



```{r}

running %>% ggplot(aes(x =Running.Time, y = "" )) + geom_boxplot() + labs(title = "Boxplot of 3k running times in sample") 

```

**10 marks**

#### Task 3: Carry out the hypothesis test and estimate a 95% confidence interval for the true mean 3k time

The code below does a two-sided t test for the null hypothesis that the true mean `Rel.14.5` is equal to 50. Adapt the code below to carry out the hypothesis test from Task 1.


```{r}
#running %>% dplyr::select(Rel.14.5) %>% t.test(mu = 50)
running %>% dplyr::select(Running.Time) %>% t.test(mu = 10.2)
```
**5 marks**

#### Task 4: Interpret (a) the results of the hypothesis test (b) the 95% confidence interval from Task 3 

Note: you need to interpret the results after you have adapted the code in Task 3.  You will not get marks for this question if you intepret the hypothesis test for the `Rel.14.5` variable.
(1-2 sentences)

(a) the results of the hypothesis test: 
The p-value of the One Sample t-test is 0.00119, which is less than the significant level 0.05. In addition, the mean of x = 9.458125, both of these factors indicate we should reject the null hypothesis



(b) the 95% confidence interval from Task 3: (9.061532 - 9.854718)
This interval shows the range of value that are 95% confident to contain the true mean. As 10.2 is > than the upper bound value of this interval, further supports the rejection of the null null hypothesis. If 10.2 is greater than the largest value in the 95% confidence interval, there is a significant chance the mean is lower than 10.2



**5 marks**

#### Task 5: A bootstrap 95% confidence interval for the mean 3k time 

The code below estimates a 90% confidence interval for the mean 3k time using the bootstrap method.
Change the number of bootstrap replications to 1000, and change the confidence level to 95%. 


```{r}
set.seed(123456789) # this is for replicable simulations - do not change this
boot <- running %>% 
          specify(response = Running.Time) %>%
          generate(reps = 1000, type = 'bootstrap') %>%
          calculate(stat = 'mean')

### Produce the 90% bootstrap confidence interval
### using the generated bootstrap distribution of the sample statistic
get_ci(boot, level = 0.95)

### plot the histogram of bootstrap statistics
### and shading the bootstrap confidence interval. 
boot %>% 
  visualize(method = "simulation") +
  shade_confidence_interval(get_ci(boot))+
  xlab("Bootstrap Mean")+
  ylab("Frequency")


```

**5 marks**

#### Task 6: Compare the 95% confidence interval using the t distribution with the bootstrap 95% CI 

(1-2 sentences)
The bootstrap 95% CI ranges from 9.096812	to 9.802625, which is very similar to the ranges found in the 95% confidence interval using the t distribution, and also does not include the value 10.2 minutes within it's range.



**5 marks**

#### Task 7: A tolerance interval for the range in which we are 95% confident 90% of the population of 3k times will lie.  

(a) The code below estimates a tolerance interval for the range in which we are 95% confident 95% of the population of 3k times will lie.  Adapt the code to estimate  a tolerance interval for the range in which we are 95% confident 90% of the population of 3k times will lie.

(b) The men's world record for the 3k is 7 mins 20.67 seconds by Daniel Komen of Kenya in 1996.  Using your adapted tolerance interval, state the running time you are 95% confident that the fastest 10% of our study population will achieve. (1 sentence) 

Running time we are  95% confident that the fastest 10% of our study population will achieve = 2-sided.lower = 7.635421

```{r}
normtol.int(running$Running.Time, alpha = 0.05, P = 0.90, side = 2)
```


**5 marks**

## Linear Regression


#### Task 8: Labelled scatterplot with smooth of `Running.Time` versus `v4mM`

(a) The code below makes a scatterplot with smooth of `Running.Time` versus `Rel.14.5`.  Adapt the code to make a scatterplot with smooth of `Running.Time` versus `v4mM`.  Remember to update the title, axis labels, and units. 

(b) What does the smoother suggest regarding the suitability of a simple linear regression model for the relationship between `Running.Time` and `v4mM`? (1 sentence)

The smoother suggests that a simple linear regression model may provide a reasonable fit for the relationship between Running.Time and v4mM, indicating that while there is a linear trend, other factors might also be influencing the running time.

```{r, warning = FALSE, message = FALSE}
running %>% ggplot(aes(x = v4mM, y = Running.Time)) + geom_point() + geom_smooth() + labs(title = "Scatterplot with smooth for 3k time versus running velocity at a blood lactate concentration of 4 mmol/L", x = "Running velocity at 4 mmol/LV (km/hr)", y = "3k time (minutes)")
```


**5 marks**

#### Task 9: Correlation coefficient between `v4mM` and `Running.Time` 

(a) The code below calculates the sample correlation coefficient between `Running.Time` and `Rel.14.5`.  Change to code to calculate the sample correlation coefficient between `Running.Time` and `v4mM`.

```{r}
running %>% dplyr::select (Running.Time, v4mM) %>% cor()
```


(b) Interpret the relationship between `Running.Time` and `v4mM`, referring to the relevant scatterplot and correlation coefficient.

The correlation coefficient = -0.925857, which is an indication of a strong negative correlation between these two variables. The Scatterplot also shows a strong negative correlation, as we see points clustered close to each other, on a line that slopes downwards from left to right.

Both of these factors indicate a  a strong and consistent inverse relationship between v4mM and Running Time.

**10 marks**
  
#### Task 10: Line of best fit for the linear relationship between 3 km running time and v4mM in the sample.  

(a) The code below calculates the simple linear regression for `Running.Time` versus `Rel.14.5`.  Change the code to produce the simple linear regression for `Running.Time` versus `v4mM`.

```{r}
lm1<-lm(Running.Time ~ v4mM, data = running)
lm1
```

**5 marks**

#### Task 11: Write down the equation of the line of best fit from for the simple linear regression of `Running.Time` on `v4mM` and provide an interpretation of the slope and intercept. Does the intercept have a physically meaningful interpretation? 
Slope: -0.3729 (change in Running Time)
Intercept: 15.8223


Equation of the line of best fit= y=mx+c = 15.8223 - 0.3729 * v4mM


The intercept does not have a physically meaningful interpretation, as it represents an estimation of Running Time when v4mM = 0 km/hr, and in a real-world scenario with well-trained athletes, it is not feasible to expect the running velocity at a blood lactate concentration of 4 mmol/L to equal zero.

**10 marks**

#### Task 12: Predict the running time (i.e. `Running.Time`) when running speed at blood lactate concentration 4 mmol/litre (i.e. `v4mM`) is 18.9 km per hour.

The code below predicts `Running.Time` when `Rel.14.5` = 48 ml/kg/min from the linear regression model of `Running.Time` on `Rel.14.5`.  It also shows the first few rows of the dataset.

(a) Adapt the code to predict `Running.Time` when `v4mM`=18.9 km/hr from a linear regression of `Running.Time` on `v4mM`.

(b) Why is the predicted running time at `v4mM`=18.9 different to 8.82, the observed running time when `v4mM` is 18.9 km/hr? - see observation 4 in the dataset.  What is the value of the residual for this observation?  



The predicted running time at `v4mM`=18.9 is different to the observed running time of 8.82 as the linear regression model, as seen above, gives an estimation based on the general trends of all the data. However, it does not consider circumstances such as individual differences and other conditions that may affect the predict the running time.

Value of the residual for this observation = Observed Running Time - Predicted Running Time = 8.82 - 8.775336 = 0.044664


(1-2 sentences)

```{r}
lm1<-lm(Running.Time ~ v4mM, data = running)
newdata <-data.frame(v4mM = 18.9)
predict(lm1, newdata = newdata)
head(running)
```


**10 marks**

#### Task 13: Predict the running time (i.e. `Running.Time`) when  `v4mM` is 2.6 km per hour. Explain if you have any concern related to this prediction.  

(a) Write code in the chunk below to predict `Running.Time` when `v4mM`= 2.6 km/hr.  (Hint: copy and adapt the relevant code from the chunk above)

(b) Do you have any concern about this prediction, and if so why? (1 sentence)

Predict the running time when `v4mM` is 2.6 km/h = 14.85286 minutes, which is quite below the expected running time for a well-trained athlete.



```{r}
lm1<-lm(Running.Time ~ v4mM, data = running)
newdata <-data.frame(v4mM = 2.6)
predict(lm1, newdata = newdata)
head(running)
```


**10 marks**

## Overall Conclusion and presentation 

#### Task 14: State your overall conclusions in relation to the two main aims of the study (3-4 sentences for each)

Aim (1) From the hypothesis test, confidence intervals for the true mean, and tolerance interval.

Given the the hypothesis test, confidence intervals for the true mean, and tolerance interval, we can reject the null hypothesis that the population mean 3k running time = 10.2 minutes. Firstly, the confidence interval does not encompass the 10.2-minute mark. In addition,  tolerance interval, which provides a range where a specified percentage of individual 3k running times are expected to fall, further reinforcing that the average performance level in this group differs from the hypothesised mean.


**5 marks**

Aim (2) From fitting a linear model for the relationship between 3k running time and the running speed at blood lactate concentration 4 mmol/litre.  

A strong linear negative correlation can be seen from fitting a linear model for the relationship between 3k running time and the running speed at blood lactate concentration 4 mmol/litre. This suggests that increases in running speed at this blood lactate concentration are associated with decreases in running times.This could suggest that v4mM is a reliable predictor of running performance, however, it may not be effective applying this linear model when working data beyond the range of data,as evidenced by predictions made for unusually low v4mM values seen in task 13.


**5 marks**

#### Task 15: Knit your .Rmd file to Word and submit the resulting Word document (alternatively knit to html, save the html as a pdf and submit the pdf).  Make sure you have edited the .Rmd to include your own name and student ID.  

If you submit an .Rmd file you will **lose 20 marks**, and if you submit a file missing your name and/or ID you will **lose 10 marks**.

